A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2022-10-26 Papers/Paper/manuscript.Rmd
Roger implemented edits

A word can be unpredictable for a variety of reasons: it could be low frequency, semantically unexpected, the start of a low-frequency syntactic construction, or a word that disambiguates prior words to a less common parse. Many targeted effects of interest are essentially looking at specific features that contribute to how predictable or unpredictable a word is. Thus incremental processing methods that are sensitive to predictability are useful for testing linguistic theories that make predictions about what words are unexpected.

@@

We followed the same process for the SPR data with results shown in Table \@ref(tab:spr-compare).  For SPR, GPT-2 and 5-gram models contain some value over each other model, which is less clear for Transformer-XL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than Transformer-XL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020, and contrasts with the Maze results, where the 5-gram model had the worst fit and did not provide additional predictive value over the other models. The fact that the neural language models, which capture deeper features of linguistic structure  [@huSystematicAssessmentSyntactic2020; @wilcox-etal:2022-using-computational-models], dominate the more superficial 5-gram models for Maze but not SPR suggests that Maze RTs reflect richer language structure-related processes during real-time comprehension than do SPR RTs.

@@

Lastly, we examined how different language models fare at predicting human RT data. We found that overall, the models were more predictive of the A-maze data than SPR data; however, the hierarchy of the model's predictive performance also differed between the A-maze and SPR datasets. This difference suggests that how well a language model predicts human RTs may depend on task. Maze RTs were by far best predicted by language models that capture rich features of natural language grammatical structure, unlike SPR RTs which were nearly equally well predicted by more superficial 5-gram models. These findings further add to the evidence that the Maze task is favorable for RT-based investigations of underlying linguistic processing in the human mind. More broadly, further comparisons between different processing methods on the same materials could be useful for a deeper understanding of how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].

>>>>

A word can be unpredictable for a variety of reasons: it could be low frequency, semantically unexpected, the start of a low-frequency syntactic construction, or a word that disambiguates prior words to a less common parse. Many targeted effects of interest can thus be potentially accomodated theoretically as specific features that influence word predictability.[^Of course, not all effects can necessarily be reduced to word predictability effects, and effects that *cannot* be reduced to word predictability may be of particular theoretical interest. Candidates include, for example, memory-based effects [@lewis-etal:2006; @levy-fedorenko-gibson:2013jml], noisy-channel error identification [@levy-etal:2009pnas], and the magnitude of processing difficulty in garden-path resolution [@van-schijndel-linzen-2021:single-stage-models; @wilcoxTargetedAssessmentIncremental2021].] Thus incremental processing methods that are sensitive to predictability are useful for testing linguistic theories that make predictions about what words are unexpected.

@@

We followed the same process for the SPR data with results shown in Table \@ref(tab:spr-compare).  For SPR, GPT-2 and 5-gram models contain some value over each other model, which is less clear for Transformer-XL and GRNN. In terms of log likelihoods, we find that GPT-2 is better than 5-gram is better than GRNN is better than Transformer-XL, although differences are small. The relatively good fit of 5-gram models to SPR data compared with neural models matches results from @huSystematicAssessmentSyntactic2020 and @wilcoxPredictivePowerNeural2020, and contrasts with the Maze results, where the 5-gram model had the worst fit and did not provide additional predictive value over the other models. While the nature of the generalizations made by these neural network-based models are not fully understood, controlled tests have suggested that their next-word predictions often reflect deeper features of linguistic structure [@warstadt-etal:2020-BLiMP; @huSystematicAssessmentSyntactic2020], such as subject--verb agreement [@marvin-linzen:2018-targeted] and wh-dependencies [@wilcox-etal:2022-using-computational-models], and are sensitive over longer context windows, than n-gram models. The fact that the neural language models, dominate the 5-gram models for Maze but not SPR thus suggests that Maze RTs may be more sensitive than SPR RTs to richer language structure-related processes during real-time comprehension.

@@

Lastly, we examined how different language models fare at predicting human RT data. We found that overall, the models were more predictive of the A-maze data than SPR data; however, the hierarchy of the model's predictive performance also differed between the A-maze and SPR datasets. This difference suggests that how well a language model predicts human RTs may depend on task. Maze RTs were by far best predicted by neural network language models, whereas SPR RTs were predicted nearly as well by 5-gram models. Our understanding of the linguistic generalization capabilities and performance of these neural network models is still limited, but controlled tests in the NLP literature that analyze their behavior on classic psycholinguistics paradigms [@linzen-etal:2016tacl; @futrell-etal:2019-neural-language-models] suggest more human-like performance than n-gram models are capable of. These findings further add to the evidence that the Maze task is favorable for RT-based investigations of underlying linguistic processing in the human mind. More broadly, further comparisons between different processing methods on the same materials could be useful for a deeper understanding of how task demands influence language processing [ex. @bartekSearchOnlineLocality2011].
